---
title: "Florida 2016 Presidential Campaign Contributions (as of October 2016)"
author: "Marty VanHoof"
output:
  html_document:
    #code_folding: hide
    toc: yes
    toc_depth: 2
  pdf_document:
    toc: yes
    toc_depth: 2
---

This report is an exploratory data analysis of campaign finance data from the Florida 2016 presidential primary elections.  A description of the different columns in the dataset is [here](https://www.dropbox.com/s/2fc15css7bp0jgq/CONTRIBUTOR_FORMAT.txt?dl=0).  I originally completed this analysis about a month before the November 2016 election.  If I was doing the analysis retrospectively, then it would probably be different.  Also, the analysis has a stream-of-consciousness style to it that is very exploratory in nature and this style doesn't often appear in polished reports.

```{r packages, message=FALSE, warning=FALSE}
# load required packages
#install.packages("genderdata", repos = "http://packages.ropensci.org", type = #"source")

library(ggplot2)
library(dplyr)
library(knitr)
library(tidyr)
library(gender)
library(gridExtra)
```

## 1. Initial Data Wrangling

In my local folder I renamed the dataset to `FL_contributions.csv`.  There is a problem with loading the dataset directly, since the row names are not unique.  To get around this, set `row.names = Null`, which reads in the first column of the dataset as a column of data rather than row names.  This will assign the first column the column name `row.names` and shift all the other column names over by one, so that now the column names are no longer associcated with the correct columns.  We can fix this by first moving all the columns over one place (assigning `x` as a name for the last column), and then finally removing this last column.  

Also, there are a bunch of columns in the dataset that are not needed:

`cand_id, memo_cd, form_tp, file_num, tran_id, contbr_st`. 

I'm just going to remove these columns.

```{r}
# load the data
fl = read.csv('FL_contributions.csv', row.names = NULL)
colnames(fl) = c(colnames(fl)[-1], "x")
fl$x = NULL

# remove the columns that we don't need
fl = subset(fl, select = -c(cand_id, memo_cd, form_tp, file_num, tran_id, contbr_st))
```

I want to create 2 more categorical variables for my analysis:  Gender and political party affiliation.  To create the variable for gender, we first split the ` contbr_nm` column into 2 new columns `FirstName` and `LastName`:

```{r}
# Create a copy of the 'fl' dataframe
fl.1 = fl

# create a new column where the names are character types rather than factors
fl.1$contbr_nm.new = as.character(fl.1$contbr_nm)

# Extract the lastname and firstname from the 'contbr_nm.new' column.  
# Create 2 new columns in the dataframe:  'LastName', 'FirstName'
fl.1 = extract(fl.1, contbr_nm.new, c('LastName', 'FirstName'), '(.*),\\s*(\\S*)\\s*.*', fill = 'left', perl = TRUE)
```

We will now use the `gender()` function in R to predict gender based on a person's firstname.

```{r}
# Use gender() function to predict gender, then merge the columns with 'fl.1'
names_gender = gender(unique(fl.1$FirstName), method = "ssa" )
names_gender = names_gender[ , c('name', 'gender')]
fl.1 = merge(fl.1, names_gender, by.x = 'FirstName', by.y = 'name', all.x = TRUE)

# Drop 'FirstName' and 'LastName' columns from the dataset and make 'gender' into a factor variable
fl.1 = subset(fl.1, select = -c(FirstName, LastName))
fl.1$gender = factor(fl.1$gender)
```



