---
title: "Florida 2016 Presidential Campaign Contributions (as of October 2016)"
author: "Marty VanHoof"
output:
  html_document:
    #code_folding: hide
    keep_md: true
    toc: yes
    toc_depth: 3
  pdf_document:
    toc: yes
    toc_depth: 3
---

This report is an exploratory data analysis of campaign finance data from the Florida 2016 presidential primary elections.  A description of the different columns in the dataset is [here](https://www.dropbox.com/s/2fc15css7bp0jgq/CONTRIBUTOR_FORMAT.txt?dl=0).  I originally completed this analysis about a month before the November 2016 election. If I was doing the analysis retrospectively, then it would probably be different.  

```{r packages, message=FALSE, warning=FALSE}
### load required packages
#install.packages("genderdata", repos = "http://packages.ropensci.org", type = #"source")

library(ggplot2)
library(dplyr)
library(knitr)
library(tidyr)
library(gender)
library(gridExtra)
```

## 1. Initial Data Wrangling

In my local folder I renamed the dataset to `FL_contributions.csv`.  There is a problem with loading the dataset directly, since the row names are not unique.  To get around this, set `row.names = Null`, which reads in the first column of the dataset as a column of data rather than row names.  This will assign the first column the column name `row.names` and shift all the other column names over by one, so that now the column names are no longer associcated with the correct columns.  We can fix this by first moving all the columns over one place (assigning `x` as a name for the last column), and then finally removing this last column.  

Also, there are a bunch of columns in the dataset that are not needed:

`cand_id, memo_cd, form_tp, file_num, tran_id, contbr_st`. 

I'm just going to remove these columns.

```{r}
### load the data
fl = read.csv('FL_contributions.csv', row.names = NULL)
colnames(fl) = c(colnames(fl)[-1], "x")
fl$x = NULL

### remove the columns that we don't need
fl = subset(fl, select = -c(cand_id, memo_cd, form_tp, file_num, tran_id, contbr_st))
```

I want to create 2 more categorical variables for my analysis:  Gender and political party affiliation.  To create the variable for gender, we first split the ` contbr_nm` column into 2 new columns `FirstName` and `LastName`:

```{r}
### Create a copy of the 'fl' dataframe
fl.1 = fl

### create a new column where the names are character types rather than factors
fl.1$contbr_nm.new = as.character(fl.1$contbr_nm)

### Extract the lastname and firstname from the 'contbr_nm.new' column.  
### Create 2 new columns in the dataframe:  'LastName', 'FirstName'
fl.1 = extract(fl.1, contbr_nm.new, c('LastName', 'FirstName'), '(.*),\\s*(\\S*)\\s*.*', fill = 'left', perl = TRUE)
```

We will now use the `gender()` function in R to predict gender based on a person's firstname.

```{r}
### Use gender() function to predict gender, then merge the columns with 'fl.1'
names_gender = gender(unique(fl.1$FirstName), method = "ssa" )
names_gender = names_gender[ , c('name', 'gender')]
fl.1 = merge(fl.1, names_gender, by.x = 'FirstName', by.y = 'name', all.x = TRUE)

### Drop 'FirstName' and 'LastName' columns from the dataset and make 'gender' into a factor variable
fl.1 = subset(fl.1, select = -c(FirstName, LastName))
fl.1$gender = factor(fl.1$gender)
```

Note that there are 9605 names where the `gender()` function returned NA:

```{r}
nrow(subset(fl.1, is.na(gender)))
```

So when we do any analysis using gender, we'll have to remember to exclude those rows from the dataset.  

Now let's create a variable for political party affiliation of the campaign contributor.  We will make the assumption that the contributor is affiliated with the same party as the candidate they donated to.

```{r}
### Create a variable for political party
democrat = c('Sanders, Bernard', 'Lessig, Lawrence', 'Clinton, Hillary Rodham',
              "O'Malley, Martin Joseph", 'Webb, James Henry Jr.')
libertarian = 'Johnson, Gary'
green = 'Stein, Jill'
other = 'McMullin, Evan'
all_cands = unique(fl.1$cand_nm)
republican = all_cands[ !(all_cands %in% c(democrat, libertarian, green, other))]

fl.1$party = with(fl.1, ifelse(cand_nm %in% democrat, 'Democrat',
                        ifelse(cand_nm %in% libertarian, 'Libertarian',
                        ifelse(cand_nm %in% green, 'Green',
                        ifelse(cand_nm %in% other, 'Other', 'Republican')))))

### Make 'party' into a factor variable
fl.1$party = factor(fl.1$party)
```

The column `contb_receipt_dt` is currently a factor variable.  Let's convert the column to date format so that we can eventually look at time series plots.

```{r}
fl.1$contb_receipt_dt = as.Date(fl.1$contb_receipt_dt, "%d-%b-%y")
```

Now we are ready to start the exploratory data analysis.

## 2. Univariate Analysis

Let's start by looking at some summaries of the dataset.  Here are the column variables in our dataset.  There are 14 of them.

```{r echo=FALSE, Univariate_Plots_1}
names(fl.1)
```

The median contribution amount for all candidates is 28.42 and the mean contribution amount is 145.97.

```{r echo=FALSE, Univariate_Plots_2}
summary(fl.1$contb_receipt_amt)
```

This suggests that the distribution of ` contb_receipt_amt` is right-skewed (and we will investigate this with some plots).  There are also negative contribution amounts which are probably refunds.

The candidate with the most individual contributions is Hillary Clinton (128353).

```{r echo=FALSE, Univariate_Plots_3}
table(fl.1$cand_nm)
```

The Democrat Party has the highest number of contributions with 211105, and the Republican Party has the 2nd highest number of contributions with 125319.

```{r echo=FALSE, Univariate_Plots_3.1}
table(fl.1$party)
```

Now we'll start some data visualization and look at the distribution of contribution amounts, omitting the negative contributions.

```{r echo=FALSE, Univariate_Plots_4}
pos_contb = subset(fl.1, contb_receipt_amt > 0)
ggplot(aes(x = contb_receipt_amt), data = pos_contb) +
  geom_histogram(binwidth = 50) 
```

As I suspected, the data is highly skewed.  The vast majority of contributions are in the smaller range, but it's really hard to see from the graph.  Let's find the 99th percentile of all contribution amounts and then limit the x-axis at this value. The 99th percentile is

```{r echo=FALSE, Univariate_Plots_5}
quantile(pos_contb$contb_receipt_amt, 0.99)
```

Now let's look at the histogram again, but with the contribution amounts limited at 2700.  We will also add some more breaks on the x and y axes in order to see the amounts better, and apply a square root transformation to the y-axis, which gives us a better view of the values in the tail of the distribution.

```{r echo=FALSE, warning = FALSE, Univariate_Plots_6}
# Omit the top 1% of contribution amounts and apply the sqrt function to the y-axis
ggplot(aes(x = contb_receipt_amt), data = pos_contb) +
  geom_histogram(binwidth = 50) +
  scale_x_continuous(limits = c(0, 2700), breaks = seq(0, 2700, 200)) +
  scale_y_sqrt(breaks = c(5000,10000,25000,50000,75000,100000)) 
```

This is much better.  We can now see that most of the contributions are $100 or less with some spikes in contributions around $300, $500, $1000, $2000, $2500.  I assume these are people making larger contributions who decided on certain nice rounded numbers to contribute.

We can plot the x values on a log10 scale in order to see the smaller contribution amounts better.  It appears that the log-transformed plot is somewhat symmetric, maybe approximately normal.

```{r echo=FALSE, warning = FALSE, Univariate_Plots_7}
ggplot(aes(x = contb_receipt_amt), data = pos_contb) +
  geom_histogram(binwidth = 0.175) +
  scale_x_log10(breaks = c(10,25,100,500,2500,10000))
```

From this plot we can easily see that a large chunk of the contributions are $100 or less, and that the mode is $25.  In fact, $100 is the 75th percentile of individual contribution amounts, so 75% of the contributions are $100 or less.

```{r echo=FALSE, warning = FALSE, Univariate_Plots_8}
quantile(pos_contb$contb_receipt_amt, 0.75)
```

I'm curious how the above distributions look when separated by gender. Are there any differences in the distributions for male versus female? 

```{r echo=FALSE, warning = FALSE, Univariate_Plots_9}
# Add a layer to the above plots to facet by gender
ggplot(aes(x = contb_receipt_amt), data = subset(pos_contb, !is.na(gender))) +
  geom_histogram(binwidth = 50) +
  scale_x_continuous(limits = c(0, 2700)) +
  scale_y_sqrt(breaks = c(5000,10000,25000,50000,75000,100000)) +
  facet_wrap(~gender)

ggplot(aes(x = contb_receipt_amt), data = subset(pos_contb, !is.na(gender))) +
  geom_histogram(binwidth = 0.175) +
  scale_x_log10(breaks = c(10,25,100,500,2500,10000)) +
  facet_wrap(~gender)
```

Okay, so it appears that the distributions are quite similar across gender, although if we summarize the contribution amounts by gender, we can see that males contribute a bit more on average.

```{r echo=FALSE, warning = FALSE, Univariate_Plots_10}
by(pos_contb$contb_receipt_amt, pos_contb$gender, summary)
```

How do the number of contributions change with respect to date? To visualize this we can look at a density plot of the variable `contb_receipt_dt` with the counts on the y-axis.

```{r echo=FALSE, warning = FALSE, Univariate_Plots_11}
# Density plot for contb_receipt date
ggplot(data = pos_contb, aes(x = contb_receipt_dt, y = ..count..)) +
  geom_density()
```

The election year in question is 2016, and this plot shows that the number of contributions increase steadily as we get closer to the election.  There are also some spikes and dips in the plot.  I want to dig deeper into 2016 and see how the contributions change from month-to-month.

```{r echo=FALSE, warning = FALSE, Univariate_Plots_11.5}
lims = as.Date(c('2016-01-01', '2016-11-01'))
ggplot(data = pos_contb, aes(x = contb_receipt_dt, y = ..count..)) +
  geom_density() +
  scale_x_date(limits = lims, date_breaks = "2 months")
```

We can see that the number of contributions increases until around March 1 and then starts to decrease again and fluctuate until about mid-May (March 1 2016 was actually Super Tuesday, and I don't know if this is related to the drop-off in contributions or not).  The contributions start to increase sharply again until July 10 or so, and then drop-off and fluctuate a bit after that.  Incidentally, July 12 was the date when Bernie Sanders endorsed Hillary Clinton (again, we don't know if these events are related or not). 

Now let's look at the number of contributions by political party.  

```{r echo=FALSE, warning = FALSE, Univariate_Plots_12}
# Create a table of mean, median, total contributions, and counts grouped by party.
# We will use the other statistics in this table in the bivariate plots section
contb_by_party = pos_contb %>% group_by(party) %>%
  summarise(mean_contb_amt = mean(contb_receipt_amt),
            median_contb_amt = median(contb_receipt_amt),
            total_contb_amt = sum(contb_receipt_amt),
            num_contb = n()) 

# Bar graph for number of contributions by party
ggplot(data = contb_by_party, aes(x = party, y = num_contb)) +
  geom_bar(stat = "identity")

select(contb_by_party, party, num_contb)
```

This graph clearly shows that the number of individual contributions to the 2 major parties (Democrat, Republican) dwarf the other contributions by far.  Also, there are many more individual contributions for Democrats (209044) than Republicans (122125).  We will see later though that the total contribution amount (sum of all contributions) is greater for Republicans than Democrats. 

Let's dig deeper into the contributions and look at the contributions by candidate. Here are the top 10 candidates by number of individual contributions.

```{r echo=FALSE, warning = FALSE, message = FALSE, Univariate_Plots_13}
# Create a table of mean, total contributions, and counts grouped by candidate
contb_by_cand = pos_contb %>% group_by(cand_nm) %>%
  summarise(mean_contb_amt = mean(contb_receipt_amt),
            total_contb_amt = sum(contb_receipt_amt),
            num_contb = n()) %>%
  arrange(desc(num_contb))

# Bar graph of top 10 num_contb by candidate
ggplot(data = contb_by_cand %>% top_n(10), aes(x = reorder(cand_nm, num_contb),
                                               y = num_contb)) +
  geom_bar(stat = 'identity') +
  coord_flip() +
  xlab('cand_nm') +
  scale_y_continuous(limits = c(0,130000), breaks = seq(0, 125000, 25000))

select(contb_by_cand %>% top_n(10), cand_nm, num_contb)
```

So we can see that Hillary Clinton has the most individual contributions (126869), followed by Bernie Sanders (81954), Donald Trump (46669), Ted Cruz (28242), etc.

Now let's use a similar bar graph to examine the top 10 occupations by number of contributions.

```{r echo=FALSE, warning = FALSE, message = FALSE, Univariate_Plots_14}
# Create a table of mean, total contributions, and counts grouped by occupation
contb_by_occ = pos_contb %>% group_by(contbr_occupation) %>%
  summarise(mean_contb_amt = mean(contb_receipt_amt),
            median_contb_amt = median(contb_receipt_amt),
            total_contb_amt = sum(contb_receipt_amt),
            num_contb = n()) %>%
  arrange(desc(num_contb))
  
contb_by_occ =
filter(contb_by_occ, contbr_occupation != 'INFORMATION REQUESTED',
         contbr_occupation != 'INFORMATION REQUESTED PER BEST EFFORTS')

# Bar graph of top 10 number of contributions by occupation
ggplot(data = contb_by_occ %>% top_n(10),
       aes(x = reorder(contbr_occupation, num_contb), y = num_contb)) +
  geom_bar(stat = 'identity') +
  coord_flip() +
  xlab('contbr_occupation')

select(contb_by_occ %>% top_n(10), contbr_occupation, num_contb)
```

It's interesting to note that retired people make by far the most individual contributions (103328), followed by people who are not employed (31766), then attornies (8055), homemakers (4653), physicians (4647), etc.

### 2.1. Discussion

There are 337402 obeservations in the `fl.1` dataset and 14 features: 
```
cmte_id, cand_nm, contbr_nm, contbr_city, contbr_zip, contbr_employer, contbr_occupation, contb_receipt_amt, contb_receipt_dt, receipt_desc, memo_text, election_tp, gender, party
```
Not all these features are investigated in the analysis.  I also decided to drop the features
```
cand_id, memo_cd, form_tp, file_num, tran_id, contbr_st
```
from the original dataset, since I didn't think I would use them. A description of the different columns in the dataset can be found [here](https://www.dropbox.com/s/2fc15css7bp0jgq/CONTRIBUTOR_FORMAT.txt?dl=0).

Most of the features are categorical variables.  The ones that are not categorical variables are the following:

* cmte_id (character)
* contbr_zip (read in as a numerical variable, but should be categorical)
* contb_receipt_amt (numerical)
* contb_receipt_dt (date)

Some other observations:

* The contribution amount with the highest frequency is $25 and 75% of the contributions are $100 or less.
* Males contribute a bit more than females on average.  The median contribution amount for females is $27 and the median contribution amount for  males is $40.  The mean contribution amount for females is $128.80 and the mean contribution amount for males is $190.40.  Since the data is so positively skewed, the median is probably a better measure of center than the mean in this case.
* Democrats have more individual contributions than Republicans (209044 for Democrat, 122125 for Republicans), but the total contribution amount for Republicans is greater than for Democrats.  This is consistent with Republicans contributing more per contribution (on average).
* Hillary Clinton has the most individual contributions (126869), followed by Bernie Sanders, Donald Trump, and Ted Cruz.
* Retired people contributed by far the most to presidential campaigns in terms of individual contributions.  This is followed by people who are not employed, then attorneys.

I created two new categorical variables in the dataset:  gender and party.  To created the gender variable, I used the `gender()` function in R to predict the gender of a contributor based on their first name.  There were 9605 names where the gender could not be determined, but this only represents about 2.8% of the dataset. I created the party variable based on which candidate the person contributed to.

The contribution amounts data is very right-skewed because the majority of contributions come in smaller amounts and there there are some bigger donors who are more sparse.  In order to have a good look at the distribution, I limited the contribution amounts at the 99th percentile and then applied a squareroot transformation to the y-axis which allows us to see the tail of the distribution better.  We can also get a good look at the smaller contribution amounts by applying a log transformation on the x-axis.

The dataset was already in a tidy format.  Some grouping operations are performed on the dataset in this section and later sections.  I created some new tables where I grouped by candidate name, occupation, and party.  I also grouped by both candidate name/date, and also gender/date.  These are useful operations to perform and makes it easier to work with some plots and see some more statistical summaries of the raw data.

In the next section I will investigate the following features in more detail:
```
cand_nm, contbr_occupation, contb_receipt_amt, contb_receipt_dt, gender, party
```
We will see that there is a dependence relationship between gender and party, and that there is also a small negative correlation between contribution date and total contribution amount.

# 3. Bivariate Analysis

Let's look at boxplots for male and female contribution amounts. We have to limit the vertical axis a lot in order to see anything, since the data is so highly skewed.

```{r echo=FALSE, warning = FALSE, Bivariate_Plots_1}
# Boxplot for contribution amounts by gender
ggplot(aes(x = gender, y = contb_receipt_amt),
       data = subset(pos_contb, !is.na(gender))) +
  geom_boxplot() + 
  coord_cartesian(ylim = c(0, 300))

by(pos_contb$contb_receipt_amt, pos_contb$gender, summary)
```

As we already observed in the previous section, males contribute a bit more than females on average, and this is relected in the fact that the boxplot for males is shifted positively relative to the boxplot for females.

Let's now look at some boxplots of contribution amounts by party.

```{r echo=FALSE, warning = FALSE, Bivariate_Plots_2}
# Make a boxplot for contribution amounts by party
ggplot(aes(x = party, y = contb_receipt_amt), data = pos_contb) +
  geom_boxplot() + 
  coord_cartesian(ylim = c(0, 500))

# summarise the contribution amounts by party
by(pos_contb$contb_receipt_amt, pos_contb$party, summary)
```

This is interesting. It seems that there are some big differences in the distributions of contribution amounts conditional on party.  The distribution of contribution amounts for "Other" seems quite skewed relative to the distributions of the other parties, but this really has no practical significance since there are only 36 individual contributions in the "Other" category.

```{r echo=FALSE, warning = FALSE, Bivariate_Plots_3}
nrow(filter(pos_contb, party == 'Other'))
```

We can also clearly see that the Democrat party has the smallest contribution amounts on average.  This is also reflected in the bar graphs given below, which show the mean and median contributions amounts by party.

```{r echo=FALSE, warning = FALSE, Bivariate_Plots_4}
# Plot conditional mean and median by party on separate bar graphs
p1 <- ggplot(data = contb_by_party, aes(x = party, y = mean_contb_amt)) +
  geom_bar(stat = "identity")

p2 <- ggplot(data = contb_by_party, aes(x = party, y = median_contb_amt)) +
  geom_bar(stat = "identity")

grid.arrange(p1, p2)

select(contb_by_party, party, mean_contb_amt, median_contb_amt)
```

The next bar graph shows the total contribution amount (sum of all contributions) by party.

```{r echo=FALSE, warning = FALSE, Bivariate_Plots_5}
# Bar graph for total contributions by party
ggplot(data = contb_by_party, aes(x = party, y = total_contb_amt)) +
  geom_bar(stat = 'identity')

select(contb_by_party, party, total_contb_amt)
```

Now we can see that the sum total of all contributions is actually greater for Republicans than Democrats even though the number of individual contributions is greater for Democrats.

Next we'll look at the top 10 candidates according to the total contribution amount.

```{r echo=FALSE, warning = FALSE, message = FALSE, Bivariate_Plots_6}
# Arrange contb_by_cand in descending order according to total_contb_amt
contb_by_cand <-contb_by_cand %>% arrange(desc(total_contb_amt))

# Bar graph of top 10 total_contb_amt by candidate
ggplot(data = contb_by_cand %>% top_n(10), aes(x=reorder(cand_nm, total_contb_amt),
                                               y = total_contb_amt)) +
  geom_bar(stat = 'identity') +
  coord_flip() +
  xlab('cand_nm')

select(contb_by_cand %>% top_n(10), cand_nm, total_contb_amt)
```